### 2020-08-25 latex/vlp.tex
Wrote the bulk of the VLP VNHC section
Missing only the figures, the energy dissipation, and a brief summary
before moving on to the simulation results

\section{Dynamics of the Variable Length Pendulum}
% Derive the dynamics of the VLP in Hamiltonian form
We will model the VLP as a point mass \(m\)
connected to a fixed pivot by a massless rod of varying length \(l\) with angle 
\(q \in \mathbb{S}^1\) from the vertical, as is seen in Figure
\ref{fig:vlp_model}. 
We will also ignore any damping and frictional forces in this model.
In a realistic VLP, the rod length \(l\) varies between some minimum
length \(\underline{l} \geq 0\) and some maximum length 
\(\overline{l} > \underline{l}\). The configuration of the VLP is the vector
\(\mathbf{q} := (q,l) \in \mathbb{S}^1 \times [\underline{l},\overline{l}]\).

\begin{figure}
   \centering
   \includestandalone[width=0.4\textwidth]{images/vlp_model}
   \caption{The representation of the variable length pendulum as a mass at the
      tip of a massless rod.}\label{fig:vlp_model}
\end{figure}

@@

\section{The VLP Constraint}
% TODO: Go through the development of the constraint and prove it gains energy
% TODO: Explain how we came up with the constraint first. Talk about the child
% on a swing, the paper showing the time-optimal control input, and describe why
% we want to remove time from the equation. Discuss the lmin and lmax choices.
% Why do we need to use a VNHC instead of a VHC: we are modelling the
% time-optimal controller in a time-independent way, which requires knowledge of
% which quadrant in (q,p)-space we're in. This means using the angle theta,
% which is inherently a VNHC.

\begin{thm}
For the variable-length pendulum, define \(\theta := \arctan_2(p,q)\). 
A VNHC of the form \(l = l(\theta)\) injects energy if there exists \(l_{avg} \in \R_{>0}\) such that
\[
   \left(l(\theta) - l_{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
\]
with the property that the inequality is strict for almost every \(\theta\).
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the average-length pendulum 
    \[
       E_{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_{avg}^2} 
                    + m g l_{avg} (1-\cos(q))
    \]

    which is non-negative and has derivative 
    \[
      \dot{E}_{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_{avg}^3\right)}
                 {l_{avg}^2l(\theta)^2}
    \]

    We will show that \(E_{avg}\) is increasing.

    % VBOX UNDERFULL WARNING HERE, IGNORE IT
    Observe that \(\sign{\sin(q)p} = \sign{\sin(2\theta)}\) and, 
    by Lemma \textbf{TODO: REF LEMMA},
    \( \sign{l(\theta)^3 - l_{avg}^3} 
     = \sign{ l(\theta) - l_{avg}}\).

     Then the derivative of \(E_{avg}\) is almost always positive, since
     \begin{align*}
        \label{eq:}
        \sign{\dot{E}_{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_{avg}^3\right)} \\
                    &= -\sign{\sin(2\theta) \left(l(\theta) - l_{avg}\right)} \\
                    &\geq 0 \text{ (by assumption)}
     \end{align*}

      Hence, \(E_{avg}\) is an anti-Lyapunov function with positive derivative,  
      so the variable-length pendulum is gaining energy.
\end{proof}

\section{Simulation Results}

>>>>

\section{Dynamics of the Variable Length Pendulum}
% Derive the dynamics of the VLP in Hamiltonian form
We will model the VLP as a point mass \(m\)
connected to a fixed pivot by a massless rod of varying length \(l\) with angle 
\(q \in \mathbb{S}^1\) from the vertical, as is seen in Figure
\ref{fig:vlp-model}. 
We will also ignore any damping and frictional forces in this model.
In a realistic VLP, the rod length \(l\) varies between some minimum
length \(\underline{l} \geq 0\) and some maximum length 
\(\overline{l} > \underline{l}\). The configuration of the VLP is the vector
\(\mathbf{q} := (q,l) \in \mathbb{S}^1 \times [\underline{l},\overline{l}]\).

\begin{figure}
   \centering
   \includestandalone[width=0.4\textwidth]{images/vlp_model}
   \caption{The representation of the variable length pendulum as a mass at the
      tip of a massless rod.}\label{fig:vlp-model}
\end{figure}

@@

\section{The VLP Constraint}
% TODO: Go through the development of the constraint and prove it gains energy
% TODO: Explain how we came up with the constraint first. Talk about the child
% on a swing, the paper showing the time-optimal control input, and describe why
% we want to remove time from the equation. Discuss the lmin and lmax choices.
% Why do we need to use a VNHC instead of a VHC: we are modelling the
% time-optimal controller in a time-independent way, which requires knowledge of
% which quadrant in (q,p)-space we're in. This means using the angle theta,
% which is inherently a VNHC.
Since we wish to define a VNHC which injects energy into the VLP in a human-like
manner,
we return once again to the example of a person standing on a swing.
As can be seen in Figure \ref{fig:child-vlp}, condensing the person 
into a center of mass gives a VLP model of their motion, where
the pendulum shrinks when the person stands and lengthens when they squat. This
is equivalent to the VLP model from Figure \ref{fig:vlp-model}.
\begin{figure}
   \centering
   \begin{subfigure}[t]{0.5\textwidth}
      % TODO: Insert child standing figure + caption
   \end{subfigure}
   \begin{subfigure}[t]{0.5\textwidth}
      % TODO: Insert child squatting figure + captions
   \end{subfigure}
   \caption{The variable length pendulum representation of a child on a
   swing}\label{fig:child-vlp}
\end{figure}

The regulation of pendulum length to add energy to the VLP is known as
``pumping"`. \citeauthor{pumping_swing_standing_squatting} asked whether
the pumping strategy performed by children is time-optimal, assuming the
children could squat or stand instantaneously
\cite{pumping_swing_standing_squatting}. Indeed, they discovered that 
a child will pump energy into the VLP as fast as is physically possible.

The pumping strategy performed by children is quite 
straightforward: they stand up at the lowest point of their swing, and squat at
the highest point. Looking at their VLP representation, the pendulum shortens
at the bottom of the swing, and lengthens at the top. 
This pumping strategy makes intuitive sense:
by shortening the pendulum at the bottom, conservation of angular momentum indicates the
mass must gain speed to compensate for the decrease in pendulum length
\cite{how_to_pump_a_swing}.
Energy is not conserved in this process, and the pendulum gains kinetic energy
which allows the pendulum to reach a higher point in its swing.
Lengthening the pendulum when it reaches this point means gravity
imparts a larger angular momentum to the mass by the time it reaches the bottom
of its swing.
By alternating these processes, the pendulum experiences a net gain in
rotational energy which is generated entirely by gravity.

The time-optimal controller described by \cite{pumping_swing_standing_squatting}
depends on a state trajectory in time. We will convert this time-based
controller into a time-independent controller which still uses this time-optimal
strategy. Notice that the control strategy depends on knowing when the
system is at the ``bottom" or ``top" of the swing. Since being at the ``bottom"
is the same as saying the pendulum angle is \(q = 0\) and being at the ``top" is
equivalent to the pendulum's momentum being \(p = 0\), our method of
VNHCs is especially well suited to this problem.

Figure \ref{fig:vlp-optimal-controller-qp-plane} shows the time-optimal strategy directly
on the \((q,p)\)-plane. The axes of this plane define where the pendulum should
change length. Since the pendulum stays the same length inside each quadrant, we
will use the angle \(\theta = \arctan_2(p,q)\) in the \((q,p)\)-plane to define
the optimal controller \(l^\star(\theta)\) (\ref{eqn:vlp-optimal-controller}).
Figure \ref{fig:vlp-optimal-controller} shows this controller as a function of
\(\theta\).
\begin{equation}\label{eqn:vlp-optimal-controller}
   l^\star(\theta):= \begin{cases}
      \underbar{l} & \theta \in [-pi, -\frac{\pi}{2}[ \cup [0,\frac{\pi}{2}[ \\
      \bar{l} & \theta \in [-\frac{\pi}{2},0[ \cup [\frac{\pi}{2}, \pi[ \\
   \end{cases}
\end{equation}

\begin{figure}
   \centering
   % TODO: Inser the image of the optimal controller in the (q,p) plane
   \caption{The time-optimal VLP controller from
   \cite{pumping_swing_standing_squatting} displayed in the
\((q,p)\)-plane.}\label{fig:vlp-optimal-controller-qp-plane}
\end{figure}

\begin{figure}
   \centering
   %TODO: Insert the image of the optimal controller as a function of theta
   \caption{The time-optimal VLP controller converted to a discontinuous
   function \(l(\theta)\).}\label{fig:vlp-optimal-controller}
\end{figure}

To find a continuously differentiable function which approximates this controller,
we can attach sinusoids at the transition points of
\(l^\star(\theta)\) (see Figure \ref{fig:vlp-T-controller}). Let
\(\Delta l := (\bar{l} - \underbar{l})/2\) and 
\(l_{\text{avg}} := (\bar{l} + \underbar{l})/2\).
Providing these newly attached sinusoids with a frequency \(\omega = \frac{pi}{T}\) 
(where \(T \in ]0,\frac{\pi}{2}]\) is a control parameter), we have a new
controller \(l_T(\theta)\):
\begin{equation}\label{eqn:vlp-T-controller}
   l_T(\theta) = \begin{cases}
      \underbar{l} & \theta \in \left[-\pi + \frac{T}{2}, -\frac{\pi}{2} - \frac{T}{2}\right] 
      \cup \left[\frac{T}{2}, \frac{\pi}{2} - \frac{T}{2}\right] \\
      \bar{l} & \theta \in \left[-\frac{\pi}{2} + \frac{T}{2}, -\frac{T}{2}\right] 
      \cup \left[\frac{\pi}{2} + \frac{T}{2}, \pi - \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta + \pi)) + l_{\text{avg}} & \theta \in
      \left[-\pi,-\pi + \frac{T}{2}\right] \\
      -\Delta l \sin(\omega(\theta - a)) + l_\text{avg} & 
      \theta \in \left[a - \frac{T}{2}, a + \frac{T}{2}\right] \text{ for } 
      a \in \left\{-\frac{\pi}{2}, 0, \frac{\pi}{2}\right\} \\
      -\Delta l \sin(\omega(\theta-\pi)) & \theta \in \left[\pi - \frac{T}{2},\pi\right] \\
   \end{cases}
\end{equation}

\begin{figure}
   \centering
   % TODO: Insert an image of the controller with sinusoids attached
   \caption{}\label{fig:vlp-T-controller}
\end{figure}

This controller approximates the optimal controller, since 
\[
   \lim\limits_{T \rightarrow 0} l_T(\theta) = l^\star(\theta)
\]
Unfortunately, while \(l_T(\theta)\) is continuously differentiable, it is not
twice continuously differentiable for most values of \(T\).
If we wish to use it as a VNHC, we must ensure that either the generalized
forces \(\tau\) acting on \(p_l\) can be discontinuous, or we must find a value
of \(T\) where this controller is smooth.
Thankfully, \(l_{\frac{\pi}{2}}(\theta)\) is smooth; it can be simplified into
the expression (\ref{eqn:vlp-smoothed-controller}), and is plotted for
demonstration in Figure \ref{fig:vlp-smoothed-controller}.
\begin{equation}\label{eqn:vlp-smoothed-controller}
   l_\frac{\pi}{2}(\theta) = -\Delta l \sin(2\theta) + l_{\text{avg}}
\end{equation}

\begin{figure}
   \centering
   % TODO: Insert the image of the smoothed VLP controller
   \caption{The smoothed VLP controller attained by setting the parameter 
      \(T\) to its upper limit
   \(\frac{\pi}{2}\).}\label{fig:vlp-smoothed-controller}
\end{figure}

We therefore set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]

This VNHC is not the same as the optimal controller
(\ref{eqn:vlp-optimal-controller}), so we need to prove it will still gain
energy. We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) of the
VLP diverge from the origin, regardless of the initial condition of the system. 
To do this, we will use the notion of an anti-Lyapunov function.

\begin{defn}\label{defn:anti-lyapunov}
   Let \(\dot{x} = f(x) \in R^n\) be an ODE with \(f(0) = 0\). 
   A positive definite function \(V(x)\) is \textbf{anti-Lyapunov} if 
   \[
      \dot{V}(x) = dV(x) f(x) \geq 0
   \]
   where the inequality is strict for almost every \(x \in \R^n\).
\end{defn}

By applying Lyapunov stability theory \cite{lyapunov},
we see that an anti-Lyapunov function
\(V\) is simply a Lyapunov function for the negative-time system 
\(\dot{x} = -f(x)\). In other words, since \(V\) proves that the orbit \(x(-t)\) 
of the system travelling backwards in time is asymptotically converging to the
equilibrium, the system must be diverging from the equilibrium in forward time.
If we can find such a function for the VLP's dynamics under our VNHC, the path
generated by the VNHC in the \((q,p)\)-plane must be diverging from the origin
so long as the initial condition is not an equilibrium.
This in turn tells us the magnitude of the momentum \(p\) is increasing every
time the path hits the \(p\)-axis, which means the VLP is, on average,
increasing in energy.

We will show that anti-Lyapunov function exists for our VNHC.
In order to do so however, we need to prove the following lemma.

\begin{lemma}\label{lemma:sign-of-cube}
   For any \(x,y \in \R\):
   \[
      \sign{x^3 - y^3} = \sign{x-y}
   \]
\end{lemma}
\begin{proof}
   Observe first that
   \[
      x^3 - y^3 =  (x-y)(x^2 + xy + y^2)
   \]
   Since \( (x-y)^2 = x^2 - 2xy + y^2) \geq 0 \), 
   \[
      \frac{x^2 + y^2}{2} \geq xy
   \]
   which means \(x^2 + xy + y^2 \geq 0\), proving the lemma.
\end{proof}

\begin{thm}\label{thm:vlp-gains-energy}
   Take the VLP with Hamiltonian dyanmics (\ref{eqn:vlp-hamiltonian}) and define
   \(\theta := \arctan_2(p,q)\). Suppose the initial condition is not one of the
   equilibria: \((q(0),p(0)) \not \in \left\{(0,0), (\pi,0)\right\}\).
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underbar{l},\bar{l}]\) such that 
   \[
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \]
   with the property that the inequality is strict for almost every \(\theta\).
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the average-length pendulum 
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    which is positive definite and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    Observe that \(\sign{\sin(q)p} = \sign{\sin(2\theta)}\) and, 
    by Lemma \ref{lemma:sign-of-cube},
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Then the derivative of \(E_\text{avg}\) is almost always positive, since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Hence, \(E_\text{avg}\) is an anti-Lyapunov function
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
\end{proof}

\begin{cor}
   The VNHC \(h(q,p) = l - l_\frac{\pi}{2}(\theta)\) injects energy into the VLP
   over time, since 
   \(\left(l_\frac{\pi}{2}(\theta) - l_\text{avg}\right)\sin(2\theta) = 
   -\Delta l \sin(2\theta) < 0\) for almost every \(\theta\).
\end{cor}

% TODO: Describe energy reduction, which comes from switching the sign

Theorem \ref{thm:vlp-gains-energy} provides an alternate explanation for why the
optimal controller \(l^\star(\theta)\) works so well at injecting energy: it
maximizes the derivative of \(E_\text{avg}\) under the restriction 
\(l(\theta) \in [\underbar{l},\bar{l}]\), so that the orbit in the
\((q,p)\)-plane diverges from the origin as fast as possible. The differentiable
controllers \(l_T(\theta)\) will also inject energy, albeit at a rate which
decreases as \(T\) increases. That said, Theorem \ref{thm:vlp-gains-energy} and the
subsequent Corollary shows that VNHCs are a powerful tool in creating effective energy
pumping strategies based on natural human motion.

% TODO: How would we stabilize a specific energy level?

\section{Simulation Results}
% TODO: Show the simulations of the VLP VNHCs (multiple of them). How does the
% energy injection time of our sin(2theta) VNHC compare to the energy injection
% time of the optimal one? How does the VNHC version compare in robustness with
% the time-based one from the pumping paper?

### 2020-08-25 latex/vnhcs.tex
Wrote the bulk of the VLP VNHC section
Missing only the figures, the energy dissipation, and a brief summary
before moving on to the simulation results

Most children have an adult pushing the swing to help them go higher, or perhaps
they are swinging on a windy day. In either case, they adjust their leg
motion accordingly when presented with these external disturbances, without
keeping track of the time it takes to kick their legs. The
standard control technique of tracking a function of time, known as
\textit{trajectory tracking}, does not work as well as human motion 
because the disturbance affecting the swing will desynchronize the leg motion
and the swing, thereby stopping the energy-gaining effects.

This is a common issue in the control of many biologically-inspired systems. 
One method which can provide more realistic, robust motion is the method of
of virtual holonomic constraints (VHCs). Instead of a robot's actuators tracking
a trajectory over time, VHCs use the actuators to enforce a relation
\(h(q) = 0\) of the configuration \cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like swimming robot
\cite{vhc_snake}.

Many authors have attempted to extend these results to enforce a relation 
\(h(q,\dot{q}) = 0\) of the full state. Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and improving bipedal locomotion \cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, this nonholonomic approach has been shown to be more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
This suggests that virtual nonholonomic constraints are more capable of
stabilizing specific energy levels while rejecting disturbances, all while
producing realistic biological motion. It is for this reason we choose to study
nonholonomic constraints in this thesis.

>>>>

% TODO: Get some sources here. Show with research that trajectory control is
% unsafe, and that path tracking is better.
Most children have an adult pushing the swing to help them go higher, or perhaps
they are swinging on a windy day. In either case, they adjust their leg
motion accordingly when presented with these external disturbances, without
keeping track of the time it takes to kick their legs. The
standard control technique of tracking a function of time, known as
\textit{trajectory tracking}, does not work as well as human motion 
because the disturbance affecting the swing will desynchronize the leg motion
and the swing, thereby stopping the energy-gaining effects. 
As has been observed in the field of mobile robotics,
a trajectory tracking controller which has lagged behind due to a disturbance
will do whatever is required to ``catch up" to
the trajectory. This can cause major safety issues, which is why mobile
roboticists prefer to design controllers which drive a vehicle towards a path in
space rather than a trajectory in time.

This path tracking approach has been extended to biologically-inspired robotics
in a method known as virtual holonomic constraints (VHCs).
Instead of a robot's actuators tracking
a trajectory over time, VHCs use the actuators to enforce a relation
\(h(q) = 0\) of the configuration \cite{vhcs_for_el_systems}. 
This method has provided incredible results in the development of 
walking robots \cite{vhc_robotic_walking, vhc_stable_walking}, 
vehicle motion \cite{vhc_bicycle, vhc_helicopter}, 
and has even been used to design a snake-like swimming robot
\cite{vhc_snake}.

The downside to VHCs is that they do not use the additional information imparted
by the generalized velocities of a system.
Many authors have attempted to extend VHCs to enforce these relations 
\(h(q,\dot{q}) = 0\) of the full state to account for this drawback. 
Since these relations use actuators to
restrict both the configuration and generalized velocity, they are called
virtual \textit{nonholonomic} constraints. This idea has been used for
human-robot interaction
\cite{vnhc_human_robot_cooperation,psd_based_vnhc_redundant_manipulator,haptic_vnhc},
error-reduction on time-delayed systems \cite{vnhc_time_delay_teleop},
and has shown marked improvements to the field of bipedal locomotion 
\cite{nhvc_dynamic_walking,
hybrid_zero_dynamics_bipedal_nhvcs,output_nhvc_bipedal_control}.
Most interestingly, the nonholonomic approach has been shown to be more robust
than standard VHCs when applied to bipedal robotics \cite{nhvc_incline_walking}.
This suggests that virtual nonholonomic constraints are more capable of
stabilizing specific energy levels while rejecting disturbances, all while
producing realistic biological motion. It is for this reason we choose to study
nonholonomic constraints in this thesis.
